Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota

### 2010-02-27 Paper.tex
Edited my sections

    \subsection{Descriptive statistics}
    In 2003, three different observers flew over the rivers. Assuming that no
    tracks are laid down between individuals' flights on the same day, all three
    people should have seen the same tracks. While there is some error in
    classifying in which plot the observer saw the actual track, each observer
    should at least be seeing the same, or nearly the same, number of sites with
    tracks. However, we did not find that to be the case. If we look only at
    days that all three people observed, there appear to be major differences
    among the proportion of sites in which the scientists detected a track
    (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among 
    the
    observers indicates that the scientists did not always correctly identify
    the tracks.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where scientists think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where scientists assume 
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer 
    can mark down the existence of a track when in reality there is none. This 
    false detection process can occur as long as there is no track, whether or 
    not the site is occupied.
    \subsection{Testing the independence assumption}
    \label{shat}
    First we considered using Markov Chains to come up with a long term
    probability of detection, which we called \(r\). We let \(p\) be the number
    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
    \(1\) as a fraction of all successive pairs that start with a \(1\). These
    probabilities can be represented by the matrix:
    \[M=\begin{bmatrix}
        p & 1-p \\
        1-q & q
    \end{bmatrix}\text{.}\]

@@

\section{Spatial correlation of the data}
One approach to dealing with the significant correlation in the probability of
observing a track in neighboring sites was to use a conditional autoregressive
(CAR) prior distribution for $\theta$, the probability that an otter lays down a
track in a site on a given day.

@@

    \subsection{Simulating spatially correlated data}
    Because we modeled spatial correlation in $\theta$, we began by simulating
    spatial correlation in the track-laying process. To do this, we added an
    additional variable to the simulated data, $\alpha$. Under the new
    simulation method, $\theta$ is the probability of an otter laying down a
    track in site $i$ given site $i$ is occupied and site $i-1$ does
    \textit{not} have a track. If site $i$ is occupied and site $i-1$ does have
    a track, then the probability of an otter laying down a track in site $i$ is
    equal to $\theta+\alpha$. This means if there is a track in one site, there
    is more likely to be a track in the next site. However, when we ran data
    simulated using this new method on the model described in section \ref{car
    model}, the model almost never converged. We tested independence of the data
    using the $\hat{s}$ statistic we developed (section \ref{shat}) and found
    that the simulated data displayed much more independence than the actual
    data. We tried simulating spatial correlation only in $\psi$ with similar
    results.

    To increase the correlation of the observed tracks we simulated data with 
    spatial
    correlation in both $\theta$ and $\psi$. Under this simulation method we
    again had the variables $\theta$ and $\alpha$, but we also had analogous
    variables $ \psi$ and $\beta$. In this case, $\psi$ is the probability an
    otter occupies site $i$ given that site $i-1$ is not occupied and $\psi+
    \beta$ is the probability that an otter occupies site $i$ given that site
    $i-1$ is occupied. With this increased spatial correlation we were able to
    more closely match the $\hat{s}$ $z$-scores found in the actual data. In
    addition, the CAR model converged more often with data simulated using this
    method, but it still did not converge consistently.

@@

>>>>

    \subsection{Descriptive statistics}
    In 2003, three different observers flew over the rivers. Assuming that no
    tracks are laid down between individuals' flights on the same day, all three
    people should have seen the same tracks. While there is some error in
    classifying in which plot the observer saw the actual track, each observer
    should at least be seeing the same, or nearly the same, number of sites with
    tracks. However, we did not find that to be the case. If we look only at
    days that all three people observed, there appear to be major differences
    among the proportion of sites in which the observers detected a track
    (Figure \ref{obsPlots}). The difference in estimated na\"ive occupancy among 
    the
    observers indicates that the observers did not always correctly identify
    the tracks.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where observers think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where observers assume 
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer 
    can mark down the existence of a track when in reality there is none. This 
    false detection process can occur as long as there is no track, whether or 
    not the site is occupied.
    \subsection{Testing the independence assumption}
    \label{shat}
    First we considered using Markov Chains to come up with a long term
    probability of detection, which we called \(r\). We let \(p\) be the number
    of \(0\)'s followed by a \(0\) as a fraction of all successive pairs that
    start with a \(0\). We let \(q\) be the number of \(1\)'s followed by a
    \(1\) as a fraction of all successive pairs that start with a \(1\). These
    probabilities can be represented by the matrix:
    \[M=\begin{bmatrix}
        p & 1-p \\
        1-q & q
    \end{bmatrix}\text{.}\]

@@

\section{Spatial correlation of the data}
One approach to dealing with the correlation in the probability of
observing a track in neighboring sites was to use a conditional autoregressive
(CAR) prior distribution for $\theta$, the probability that an otter lays down a
track in a site on a given day.

@@

    \subsection{Simulating spatially correlated data}
    Because we modeled spatial correlation in $\theta$, we began by simulating
    spatial correlation in the track-laying process. To do this, we added an
    additional variable to the simulated data, $\alpha$. Under the new
    simulation method, $\theta$ is the probability of an otter laying down a
    track in site $i$ given site $i$ is occupied and site $i-1$ does
    \textit{not} have a track. If site $i$ is occupied and site $i-1$ does have
    a track, then the probability of an otter laying down a track in site $i$ is
    equal to $\theta+\alpha$. This means if there is a track in one site, there
    is more likely to be a track in the next site. However, when we ran data
    simulated using this new method on the model described in section \ref{car
    model}, the model almost never converged. We tested independence of the data
    using the $\hat{s}$ statistic we developed (section \ref{shat}) and found
    that the simulated data displayed much more independence than the actual
    data. We tried modeling spatial correlation in the occupancy process, $\psi
    $, instead of in the track-laying process, but the simulated data were again 
    more independent than the real data.

    To increase the correlation between sites we simulated data
    correlated in both the track-laying and occupancy processes. Under this 
    simulation method we
    again had the variables $\theta$ and $\alpha$, but we also had analogous
    variables $ \psi$ and $\beta$. In this case, $\psi$ is the probability an
    otter occupies site $i$ given that site $i-1$ is not occupied and $\psi+
    \beta$ is the probability that an otter occupies site $i$ given that site
    $i-1$ is occupied. With this increased spatial correlation we were able to
    more closely match the $\hat{s}$ $z$-scores found in the actual data. In
    addition, the CAR model converged more often with data simulated using this
    method, but it still did not converge consistently.

@@

\section{Acknowledgments}
We would like to thank our adviser, Bob Dobrow, for all of his help and advice and John Fieberg and John Erb from the Minnesota Department of Natural Resources for collecting and sharing the data as well as for offering their specific knowledge of the subject.
